---
title: "Impact of 'walk around flocks' CBSM campaigns on PIPL fledging success"
author: "Sarah Saunders, National Audubon Society"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

This analysis examines whether community-based social marketing (CBSM) campaigns encouraging people to "walk around flocks" positively influence piping plover (*Charadrius melodus*) fledging success at four sites in Connecticut. The analysis follows these steps:

1. **Data Preparation** - Read, clean, and filter productivity data.
2. **Mapping** - Visualize fledging success at nest locations along the Connecticut coast.
3. **Modeling** - Fit a Generalized Additive Model (GAM) to analyze predictors of fledging success.
4. **Visualization** - Generate plots of model predictions.

```{r setup, include=FALSE}
library(lubridate)
library(ggplot2)
library(dplyr)
library(tidyr)
library(mgcv)
library(sf)
library(patchwork)
library(rnaturalearth)
library(cowplot)
library(pROC)
```

## Set working directory to the location of the RMarkdown file and data files
```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

## Read and clean productivity data
```{r}
prod_atl <- read.csv("Phase4_Productivity_Jan2025.csv", header = TRUE)
prod_pipl <- prod_atl %>% filter(Species == "PIPL")

# Correct erroneous egg count entry
prod_pipl[73, "Eggs_hatch"] <- 4

prod_pipl <- prod_pipl %>%
  mutate(Hatch_success = Eggs_hatch / Eggs_laid,
         Fledge_success = ifelse(Eggs_hatch > 0, Fledglings / Eggs_hatch, NA)) %>%
  filter(!is.na(Fledge_success))

table(prod_pipl$Campaign)
```

## Convert dates and prep for analysis
```{r}
prod_pipl_beta <- prod_pipl %>%
  mutate(Fledge_success = (Fledge_success * (n() - 1) + 0.5) / n(),
         State=factor(State),
         Site=factor(Site),
         Campaign=factor(Campaign),
         Species=factor(Species),
         Exclosed=factor(Exclosed), 
         Fencing=factor(Fencing), 
         Area_closed=factor(Area_closed), 
         Brood_fate=factor(Brood_fate))  

#Correct one unknown hatch date
prod_pipl_beta <- prod_pipl_beta %>%
  mutate(Date_hatch = ifelse(Date_hatch == "U", "7/01/2024", Date_hatch))
prod_pipl_beta$Date_hatch <- mdy(prod_pipl_beta$Date_hatch)
prod_pipl_beta$JDate_hatch <- yday(prod_pipl_beta$Date_hatch)

summary(prod_pipl_beta)
```

# Mapping fledging success for visualization of spatial variation
```{r, echo=FALSE, warning=FALSE}
prod_pipl_sf <- st_as_sf(prod_pipl_beta, coords = c("Longitude", "Latitude"), crs = 4326)
ct_map <- ne_states(country = "United States of America", returnclass = "sf") %>%
  filter(name == "Connecticut")
bbox_zoom <- st_bbox(prod_pipl_sf) + c(-0.1, -0.1, 0.1, 0.1)  # Expands bbox slightly for better framing

main_map <- ggplot() +
  geom_sf(data = ct_map, fill = "gray90", color = "black") +  
  geom_jitter(data = prod_pipl_beta, aes(x = Longitude, y = Latitude, fill = Fledge_success, shape = Campaign),
              width = 0.05, height = 0.01, size = 3, stroke = 0.5, alpha = 0.8, color = "black") +  
  scale_fill_gradient(low = "#E0F3DB", high = "#08589E", name = "Fledging Success") +
  scale_shape_manual(values = c(21, 24), name = "Campaign Type") +  
  coord_sf(xlim = c(bbox_zoom["xmin"], bbox_zoom["xmax"]), 
           ylim = c(bbox_zoom["ymin"], bbox_zoom["ymax"])) +  
  theme_minimal() +
  theme(axis.text = element_blank(),  
        axis.ticks = element_blank(),  
        panel.grid = element_blank(),  
        axis.title = element_blank(),
        legend.position = "none",  
        panel.border = element_rect(fill = NA, color = "black", linewidth = 1))

fill_legend <- cowplot::get_legend(
  ggplot() +
    geom_point(data=prod_pipl_beta,aes(x = 1, y = 1, fill = Fledge_success), size = 3) +
    scale_fill_gradient(low = "#E0F3DB", high = "#08589E", name = "Fledging Success") +
    theme_minimal() +
    theme(legend.position = "right")  # Keep vertical
)

shape_legend <- cowplot::get_legend(
  ggplot() +
    geom_point(aes(x = 1, y = 1, shape = factor(c("None", "Walkers"))), size = 5) +
    scale_shape_manual(values = c(21, 24), name = "Campaign Type") +
    theme_minimal() +
    theme(legend.position = "right")  
)

inset_map <- ggplot() +
  geom_sf(data = ct_map, fill = "white", color = "black") +  
  geom_rect(aes(xmin = bbox_zoom["xmin"], xmax = bbox_zoom["xmax"], 
                ymin = bbox_zoom["ymin"], ymax = bbox_zoom["ymax"]), 
            color = "gold", fill = NA, size = 1) +  
  theme_void() +  
  theme(panel.border = element_rect(fill = NA, color = "black", linewidth = 0.5))

final_map <- ggdraw() +
  draw_plot(main_map) +
  draw_plot(inset_map, x = 0.03, y = 0.55, width = 0.2, height = 0.2) +  
  draw_plot(shape_legend, x=0.6, y=0.2, width=0.2, height=0.3)+
  draw_plot(fill_legend, x = 0.78, y = 0.3, width = 0.2, height = 0.2)  

final_map
```

# GAM model
## Goal: Assess whether ‘walk around flocks’ campaigns influence PIPL fledging success
```{r}
fledge_mod <- gam(Fledge_success ~ Campaign + s(JDate_hatch) + 
                    s(Longitude, Latitude, k=5), 
                  data = prod_pipl_beta,
                  family = betar(link = "logit"),  
                  method = "REML")  

summary(fledge_mod)
```

## Model fit assessment
```{r, warning=FALSE}
prod_pipl_beta$predicted <- predict(fledge_mod, type = "response")
prod_pipl_beta$observed <- ifelse(prod_pipl_beta$Fledge_success > median(prod_pipl_beta$Fledge_success, na.rm = TRUE), 1, 0)
auc_value <- auc(prod_pipl_beta$observed, prod_pipl_beta$predicted)
print(paste("AUC:", auc_value))
```

# Visualizations

## Key finding #1: 'Walk around flocks' campaign is positively related to fledging success.
Visual: Predicted fledging success by campaign type ('walk around flocks' vs no campaign)
```{r, echo=FALSE}
new_data_campaign <- expand.grid(
  Campaign = unique(prod_pipl_beta$Campaign),
  JDate_hatch = mean(prod_pipl_beta$JDate_hatch, na.rm = TRUE), #use average hatch date
  Longitude = mean(prod_pipl_beta$Longitude, na.rm = TRUE),
  Latitude = mean(prod_pipl_beta$Latitude, na.rm = TRUE)
)
pred_campaign <- predict(fledge_mod, newdata = new_data_campaign, type = "response", se.fit = TRUE)
new_data_campaign$Fledge_fit <- pred_campaign$fit
new_data_campaign$Fledge_se <- pred_campaign$se.fit

new_data_campaign <- new_data_campaign %>%
  mutate(Fledge_se80 = 1.28 * Fledge_se,  # 80% CI multiplier
         Fledge_se95 = 1.96 * Fledge_se)  # 95% CI multiplier
ggplot(new_data_campaign, aes(x = Campaign, y = Fledge_fit)) +
  geom_errorbar(aes(ymin = Fledge_fit - Fledge_se95, ymax = Fledge_fit + Fledge_se95), 
                width = 0, color = "darkblue", size = 1, alpha = 0.5) +
  geom_errorbar(aes(ymin = Fledge_fit - Fledge_se80, ymax = Fledge_fit + Fledge_se80), 
                width = 0, color = "darkblue", size = 2, alpha = 0.5) +
  geom_point(size = 4, color = "darkblue") +
  labs(y = "Predicted Fledging Success", x = "Campaign") +
  theme_minimal(base_size = 14)

```

## Key finding #2: Fledging success is highest with mid-season hatch dates
Visual: Predicted fledging success by hatch date (Julian)
```{r, echo=FALSE}
jdate_range <- range(prod_pipl_beta$JDate_hatch, na.rm = TRUE)
jdate_90 <- quantile(prod_pipl_beta$JDate_hatch, probs = c(0.05, 0.95), na.rm = TRUE)

new_data_jdate <- expand.grid(
  Campaign = levels(prod_pipl_beta$Campaign)[2],  # Use the walker campaign level since more obs
  JDate_hatch = seq(jdate_90[1], jdate_90[2], length.out = 100),
  Longitude = mean(prod_pipl_beta$Longitude, na.rm = TRUE),
  Latitude = mean(prod_pipl_beta$Latitude, na.rm = TRUE)
)
pred_jdate <- predict(fledge_mod, newdata = new_data_jdate, type = "response", se.fit = TRUE)
new_data_jdate$Fledge_fit <- pred_jdate$fit
new_data_jdate$Fledge_se <- pred_jdate$se.fit

ggplot(new_data_jdate, aes(x = JDate_hatch, y = Fledge_fit)) +
  geom_line(size = 1.2, color = "darkblue") +
  geom_ribbon(aes(ymin = Fledge_fit - 1.96 * Fledge_se, ymax = Fledge_fit + 1.96 * Fledge_se), 
              alpha = 0.2, fill = "darkblue") +
  labs(y = "Predicted Fledging Success", x = "Hatch Date (Julian)") +
  theme_minimal(base_size = 14) 

```

# Conclusion
This analysis suggests that 'walk around flocks' campaigns may have a positive effect on piping plover fledging success. However, further studies incorporating larger datasets and site-specific environmental factors are needed to strengthen these findings.

# Key takeaways
1. ‘Walk around flocks’ campaign positive effect is suggestive of efficacy in improving PIPL fledging success, but can’t definitively conclude an impact with limited data.

2. Nests hatching mid-season are more successful.

3. Other site-specific factors likely influencing success due to significant effect of spatial variation.
